### 2013-07-11 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check on nonlinear chapter

We therefore devise a new nonlinear method based on the MCSA algorithm
and Newton's method. In this chapter the fundamentals of Newton's
method will be presented along with a brief discussion of inexact
Newton methods. Newton-Krylov methods, a variant of inexact Newton
methods, will be outlined along with two common techniques for
Jacobian generation. Following this, the new algorithm will be
presented and implementation details will be discussed. Using the new
algorithm and a production fluid physics code, a group of nonlinear
benchmark problems will be used to verify the correctness of the
method relative to solutions obtained using conventional
methods. Finally, will compare the performance of the new method
against contemporary Newton-Krylov methods for the benchmark problems.

@@

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver. Furthermore, they bind the convergence of the outer nonlinear
iteration to the inner linear iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desireable direction such that convergence properties can be
  improved when the initial guess for $\ve{u}$ is poor
  \citep{pawlowski_globalization_2006}.

@@

\subsection{Parallel FANM Implementation}
\label{subsec:parallel_fanm}
Here we briefly comment on how a FANM method may be implemented in
parallel, defering the bulk of the discussion on parallel methods
Monte Carlo to Chapter~\ref{ch:parallel_methods}. A parallel FANM
method relies on a basic set of parallel matrix-vector operations
outlined in \S~\ref{sec:parallel_krylov_methods} as well as the global
residual and Jacobian assembly procedure described in
\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
system in parallel through the element-wise function evaluations to
generate both the global Jacobian operator and the global residual
vector on the right hand side. Per Bartlett's work, efficient and
automated parallel mechanisms are available to do this through a
sequence of scatter/gather operations. With these tools available for
residual and Jacobian generation, the remainder of the parallel
procedure is simple. The linear Newton correction system is solved
using the parallel MCSA method that will be described in
Chapter~\ref{ch:parallel_methods} and the Newton correction applied to
the previous iterate's solution through a parallel vector update.

@@

\subsection{Backward-Facing Step Problem}
\label{subsec:backward_facing_step}
The third benchmark was generated by Gartling in 1990 and consists of
both flow over a backward step and an outflow boundary condition
\citep{gartling_test_1990}. Using the mass and momentum equations
while neglecting the energy equation, this problem utilizes a longer
domain with a 1/30 aspect ratio with the boundary conditions as shown
in Figure~\ref{fig:backward_facing_step}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.3}{
      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the backward facing step
      benchmark.} \textit{Zero velocity boundary conditions are
      applied at the top and bottom of the domain while the outflow
      boundary condition on the right boundary is represented by zero
      stress tensor components in the direction of the flow. For the
      inlet conditions, the left boundary is split such that the top
      half has a fully formed parabolic flow profile and the bottom
      half has a zero velocity condition, simulating flow over a
      step.}}
  \label{fig:backward_facing_step}
\end{figure}
In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. The flow over this step will generate a recirculating backflow
under the inlet flow towards the step. As in the lid driven cavity
problem, the nonlinear behavior of this benchmark and the difficulty
in obtaining a solution is dictated by the Reynolds number of the
fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
was used to discretize the domain with Reynolds number up to
\sn{8}{2}.

>>>>

We therefore devise a new nonlinear method based on the MCSA algorithm
and Newton's method. In this chapter the fundamentals of Newton's
method will be presented along with a brief discussion of inexact
Newton methods. Newton-Krylov methods, a variant of inexact Newton
methods, will be outlined along with two common techniques for
Jacobian generation. Following this, the new algorithm will be
presented and implementation details will be discussed. Using the new
algorithm and a production fluid physics code, a group of nonlinear
benchmark problems will be used to verify the correctness of the
method relative to solutions obtained using conventional methods in
various flow regimes. Finally, we will compare the performance of the
new method against contemporary Newton-Krylov methods for the same
benchmark problems.

@@

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver. Furthermore, they bind the convergence of the outer nonlinear
iteration to the inner linear iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that convergence properties can be
  improved when the initial guess for $\ve{u}$ is poor
  \citep{pawlowski_globalization_2006}.

@@

\subsection{Parallel FANM Implementation}
\label{subsec:parallel_fanm}
Here we briefly comment on how a FANM method may be implemented in
parallel, deferring the bulk of the discussion on parallel methods
Monte Carlo to Chapter~\ref{ch:parallel_methods}. A parallel FANM
method relies on a basic set of parallel matrix-vector operations
outlined in \S~\ref{sec:parallel_krylov_methods} as well as the global
residual and Jacobian assembly procedure described in
\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
system in parallel through the element-wise function evaluations to
generate both the global Jacobian operator and the global residual
vector on the right hand side. Per Bartlett's work, efficient and
automated parallel mechanisms are available to do this through a
sequence of scatter/gather operations. With these tools available for
residual and Jacobian generation, the remainder of the parallel
procedure is simple. The linear Newton correction system is solved
using the parallel MCSA method that will be described in
Chapter~\ref{ch:parallel_methods} and the Newton correction applied to
the previous iterate's solution through a parallel vector update.

@@

\subsection{Backward-Facing Step Problem}
\label{subsec:backward_facing_step}
The third benchmark was generated by Gartling in 1990 and consists of
both flow over a backward step and an outflow boundary condition
\citep{gartling_test_1990}. Using the mass and momentum equations
while neglecting the energy equation, this problem utilizes a longer
domain with a 1/30 aspect ratio with the boundary conditions as shown
in Figure~\ref{fig:backward_facing_step}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.3}{
      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the backward facing step
      benchmark.} \textit{Zero velocity boundary conditions are
      applied at the top and bottom of the domain while the outflow
      boundary condition on the right boundary is represented by zero
      stress tensor components in the direction of the flow. For the
      inlet conditions, the left boundary is split such that the top
      half has a fully formed parabolic flow profile and the bottom
      half has a zero velocity condition, simulating flow over a
      step.}}
  \label{fig:backward_facing_step}
\end{figure}
In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. The flow over this step will generate a recirculating back-flow
under the inlet flow towards the step. As in the lid driven cavity
problem, the nonlinear behavior of this benchmark and the difficulty
in obtaining a solution is dictated by the Reynolds number of the
fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
was used to discretize the domain with Reynolds number up to
\sn{8}{2}.

### 2013-07-11 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check on nonlinear chapter

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative performance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. Furthermore, no overlap was required to maintain
iterative performance be letting histories near the boundary take a
few more steps in their random walks. If in the future there are
ill-conditioned problems where using subdomain Monte Carlo reduces the
algorithmic efficiency of MCSA due to the information lost in the
correction vector, overlap would be a viable mechanism to potentially
maintain that iterative performance, keeping in mind the potential
impacts on scalability that were observed in this work.

>>>>

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative performance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. Furthermore, no overlap was required to maintain
iterative performance be letting histories near the boundary take a
few more steps in their random walks. If in the future there are
ill-conditioned problems where using subdomain Monte Carlo reduces the
algorithmic efficiency of MCSA due to the information lost in the
correction vector, overlap would be a viable mechanism to potentially
maintain that iterative performance, keeping in mind the potential
impacts on scalability that were observed in this work. However,
although eliminating all Monte Carlo communication was observed to be
the best mode of use for MCSA, the parallel Monte Carlo methods
developed for this work still provide value for cases where only a
Monte Carlo solution is desired. In those cases, full transport of the
Monte Carlo histories are required in order to generate the correct
solution.
